---
title: "Analysis of Capital Asset Pricing Model (CAPM)"
author: 
  - "Filip Farkašovský"
  - "(Skupina 02, Tím 08)"
date: "`r format(Sys.Date(), '%d. %m. %Y')`"
lang: eng
output:
  bookdown::pdf_document2:
    highlight: monochrome
    fig_caption: true
---

# Abstract {-}

This study examines the Capital Asset Pricing Model (CAPM) using historical stock data from S&P 500 constituents. The analysis focuses on estimating beta coefficients for individual assets and evaluating whether systematic risk adequately explains differences in expected returns.

# Data Collection and Estimation of Beta Coefficients

## Data Loading and Preparation

```{r setup, include=FALSE}
rm(list = ls())
set.seed(1789)


knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
  fig.width = 5,
  fig.height = 3,
    out.width = "60%",
  fig.align = "center")

library(knitr)
library(tidyverse)
library(lubridate)
library(quantmod)
library(PerformanceAnalytics)
library(broom)
library(dplyr)
library(ggplot2)
library(car)
library(kableExtra)
```

```{r data-loading, include=FALSE}
rm(list = ls())

# Load and preprocess stock dataset
unzip("SP500_Historical_Data.zip")
stocks <- read_csv("SP500_Historical_Data.csv") %>% 
    mutate(ticker = factor(Ticker),
           date = as.Date(Date),
           close = as.numeric(Close)) %>%
    arrange(ticker, date)

# Monthly prices
stocks <- stocks %>%
  mutate(year_month = floor_date(date, "month")) %>%
  arrange(ticker, year_month) %>%
  group_by(ticker, year_month) %>%
  summarise(price = last(close[!is.na(close)]), .groups = "drop") 

# Load and preprocess DTB3 dataset
risk_free_rate <- read_csv("DTB3.csv") %>%
  mutate(
    date = as.Date(observation_date),
    rf_annual = as.numeric(DTB3)
  ) %>%
  mutate(year_month = floor_date(date, "month")) %>%
  group_by(year_month) %>%
  summarise(rf_annual = last(rf_annual[!is.na(rf_annual)]), .groups = "drop") %>%
  mutate(rf_monthly = (rf_annual / 100) / 12)

# Join datasets and filter stock with few samples
capm_data <- stocks %>%
  left_join(risk_free_rate, by = "year_month")

n_before <- capm_data %>%
  summarise(n_tickers = n_distinct(ticker)) %>%
  pull(n_tickers)

# Filter stocks with insufficient number of observations
capm_data <- capm_data %>%
  group_by(ticker) %>%
  filter(!is.na(price)) %>%
  filter(n() >= 60) %>%
  ungroup()

n_after <- capm_data %>%
  summarise(n_tickers = n_distinct(ticker)) %>%
  pull(n_tickers)

# Logaritmical monthly revenue
capm_data <- capm_data %>%
  arrange(ticker, year_month) %>%
  group_by(ticker) %>%
  mutate(stock_return = log(price / lag(price))) %>%
  filter(!is.na(stock_return)) %>%
  ungroup()

# Obtain market returns for equal weighted stock portfolio
market_returns <- capm_data %>%
  group_by(year_month) %>%
  summarise(market_return = mean(stock_return), .groups = "drop")

capm_data <- capm_data %>%
  left_join(market_returns, by = "year_month")

# Add excess returns 
capm_data <- capm_data %>%
  mutate(
    excess_return = stock_return - rf_monthly,
    excess_market = market_return - rf_monthly
  )
```

We use a dataset containing historical daily stock price data for 472 current S&P 500 companies, covering the period from January 2000 to February 2026. This dataset is one of the most comprehensive freely available S&P 500 datasets on Kaggle, consisting of approximately 2.7 million observations over 26 years of U.S. equity market history. The dataset is available at: \url{https://www.kaggle.com/datasets/jacksaleeby/s-and-p500-historical-data}. To ensure sufficient time series length for reliable beta estimation, we retain only stocks with at least 60 monthly observations. After applying this filter, the dataset is reduced from `r n_before` stocks to `r n_after` stocks.

Monthly returns are computed from end-of-month prices as logarithic returns:

\begin{equation}
r_{j,t} = \ln\left(\frac{P_{j,t}}{P_{j,t-1}}\right)
\end{equation}

As the risk-free rate, we use the 3-Month Treasury Bill Secondary Market Rate (Discount Basis), which is a commonly used proxy for the risk-free rate in econometric studies. The data are obtained from the Federal Reserve Economic Data (FRED) database: \url{https://fred.stlouisfed.org/series/DTB3}.

The series is available at a daily frequency and in annualized percentage terms. It is converted to monthly frequency by taking the last observation of each month and transforming it into a monthly rate:

\begin{equation}
r_{f,t} = \frac{r_{f,t}^{(annual)}}{100 \cdot 12}.
\end{equation}

This transformation provides an approximation of the effective monthly risk-free rate.

## Visualization of Time Series
The time series in Figure @ref(fig:visualization) shows the evolution of the market return and the risk-free rate over time. The risk-free rate remains close to zero for most of the sample period and exhibits only limited variation, while the market return shows substantially higher volatility.


```{r visualization, fig.width=5, fig.height=3, out.width="70%", fig.cap="Time Series of Market Return and Risk-Free Rate"}
capm_data %>%
    ggplot(aes(x = year_month)) +
  geom_line(aes(y = market_return, color = "Market return"), linewidth = 0.8) +
  geom_line(aes(y = rf_monthly, color = "Risk-free rate"), linewidth = 0.8) +
  scale_color_manual(values = c("Market return" = "lightskyblue",
                                "Risk-free rate" = "tomato")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::percent_format()) +

  labs(
    x = "Year",
    y = "Monthly return",
    color = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Table @ref(tab:summary-stat) reports summary statistics of monthly stock returns. The mean return is close to zero, while the dispersion is substantial, as indicated by the standard deviation.

```{r summary-stat}
capm_data %>%
  summarise(
    Mean = mean(stock_return, na.rm = TRUE),
    SD = sd(stock_return, na.rm = TRUE),
    Min = min(stock_return, na.rm = TRUE),
    Q1 = quantile(stock_return, 0.25, na.rm = TRUE),
    Median = median(stock_return, na.rm = TRUE),
    Q3 = quantile(stock_return, 0.75, na.rm = TRUE),
    Max = max(stock_return, na.rm = TRUE)
  ) %>%
  kable(digits = 4,booktabs = TRUE,
 caption = "Summary statistics of monthly stock returns")
```

Figure @ref(fig:stocks-dist) presents the distribution of monthly returns for a sample of stocks.

```{r stocks-dist, fig.cap="Distribution of month returns - 10 sample stocks"}
set.seed(1789)
sample_stocks <- sample(unique(capm_data$ticker), 10)


capm_data %>%
  filter(ticker %in% sample_stocks) %>%
  mutate(ticker = fct_reorder(ticker, stock_return, median)) %>%
  ggplot(aes(x = ticker, y = stock_return)) +
  geom_boxplot(fill = "lightskyblue", alpha = 0.6, outlier.size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "",
    x = NULL, y = "Month return"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```

## Estimate of betta coefficients of the characteristic curves

For each stock, the coefficients of the characteristic curve are estimated using the following model: 
\begin{equation}
r_{j} - r_{f} = \alpha_{j} + \beta_{j} (r_{m} - r_{f}) + \varepsilon_{j}
\end{equation}

```{r coeficient-estimation}
coef_characteristic_curve <- capm_data %>%
  group_by(ticker) %>%
  do(tidy(lm(excess_return ~ excess_market, data = .))) %>%
  ungroup() %>%
  filter(term %in% c("(Intercept)", "excess_market")) %>%
  select(ticker, term, estimate, std.error, statistic, p.value) %>%
  pivot_wider(names_from = term, 
              values_from = c(estimate, std.error, statistic, p.value)) %>%
  rename(
    alpha     = `estimate_(Intercept)`,
    beta      = estimate_excess_market,
    alpha_se  = `std.error_(Intercept)`,
    beta_se   = std.error_excess_market,
    alpha_t   = `statistic_(Intercept)`,
    beta_t    = statistic_excess_market,
    alpha_p   = `p.value_(Intercept)`,
    beta_p    = p.value_excess_market
  )

n_defensive_betas <- sum(coef_characteristic_curve$beta < 1)
n_agressive_betas <- sum(coef_characteristic_curve$beta > 1)
```

The CAPM beta coefficients are obtained from time-series regressions of excess stock returns on excess market returns. The results show substantial heterogeneity in systematic risk across assets.

As reported in Table \@ref(tab:tab-beta-summary), the average $\beta$ is close to one, which suggests that, on average, stocks have similar sensitivity to market movements as the market portfolio. However, individual estimates vary considerably. Some stocks have $\beta > 1$, indicating higher systematic risk, while others have $\beta < 1$ or values close to zero.

Table \@ref(tab:tab-beta-summary) also reports the proportion of statistically significant coefficients at different significance levels (5%, respectively 1%). These values represent the percentage of stocks for which the null hypothesis $\beta = 0$ is rejected at the corresponding level. A higher percentage implies stronger overall evidence that market returns explain individual stock returns.

The Security Characteristic Lines of selected stocks are shown in Figure \@ref(fig:scl-plot). 

```{r tab-beta-summary}
coef_characteristic_curve %>%
  summarise(
    Mean = mean(beta, na.rm = TRUE),
    Median = median(beta, na.rm = TRUE),
    Minimum = min(beta, na.rm = TRUE),
    Maximum = max(beta, na.rm = TRUE),
    `Significant (5%) [%]`  = mean(beta_p < 0.05, na.rm = TRUE) * 100,
    `Significant (1%) [%]`  = mean(beta_p < 0.01, na.rm = TRUE) * 100
  ) %>%
  kable(
    digits = 3,
    caption = "Summary statistics of estimated CAPM beta coefficients",
    booktabs = TRUE,
    align = "c"
  )
```

```{r scl-plot, fig.cap="Security Characteristic Lines of selected stocks"}

set.seed(123)

sample_stocks <- coef_characteristic_curve %>%
  slice_sample(n = 6) %>%
  pull(ticker)

capm_data %>%
  filter(ticker %in% sample_stocks) %>%
  ggplot(aes(x = excess_market, y = excess_return)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "#d62728") +
  facet_wrap(~ticker, scales = "free") +
  labs(
    x = "Excess market return",
    y = "Excess stock return"
  ) +
  theme_minimal()
```


## Testing hypotheses on the intercept of the CAPM model

For each stock, we test the hypothesis that the intercept (alpha) in the CAPM model is equal to zero:

\begin{equation}
H_0: \alpha_j = 0 \quad \text{vs.} \quad H_1: \alpha_j \neq 0
\end{equation}

The tests are conducted at the 5% significance level using t-tests based on the estimated regression coefficients.

The results are summarized in Table \@ref(tab:alpha-test-summary). The null hypothesis is rejected for a relatively small subset of stocks. For these few assets, the intercept differs significantly from zero, suggesting that the CAPM does not fully explain their returns.

Most stocks do not have an intercept that is significantly different from zero. This means that, for most assets, there is no evidence that they generate returns beside what would be expected after accounting for market risk.


```{r alpha-test-summary, out.width="70%"}
library(gt)

coef_characteristic_curve %>%
  mutate(`Significant on 5%` = alpha_p < 0.05) %>%
  group_by(`Significant on 5%`) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  kable(
    digits = 2,
    booktabs = TRUE,
    align = "c",
    caption = "CAPM intercept significance (5\\% level)"
  )
```


# Testing CAPM Validity

## Portfolio Formation and Beta Aggregation

Stocks are sorted by their estimated $\beta$ coefficients and assigned to 20 portfolios of roughly equal size. Each portfolio is equally weighted, so the portfolio beta is the arithmetic mean of individual betas. Portfolio excess returns are computed as the time-series average of equally weighted stock excess returns over the full sample period.
```{r portfolio-formation}
n_portfolios <- 20

portfolio_assignments <- coef_characteristic_curve %>%
  arrange(beta) %>%
  mutate(portfolio = ntile(beta, n_portfolios))

portfolio_returns <- capm_data %>%
  left_join(portfolio_assignments %>% select(ticker, portfolio), by = "ticker") %>%
  filter(!is.na(portfolio)) %>%
  group_by(portfolio, year_month) %>%
  summarise(
    portfolio_excess_return = mean(excess_return, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(portfolio) %>%
  summarise(
    mean_excess_return = mean(portfolio_excess_return, na.rm = TRUE),
    .groups = "drop"
  )

portfolio_betas <- portfolio_assignments %>%
  group_by(portfolio) %>%
  summarise(
    portfolio_beta = mean(beta, na.rm = TRUE),
    n_stocks = n(),
    .groups = "drop"
  )

portfolio_data <- portfolio_betas %>%
  left_join(portfolio_returns, by = "portfolio")
```

Figure @ref(fig:sml-plot) reveals a pattern inconsistent with CAPM. While portfolios are clearly ordered by beta, mean excess returns do not increase systematically with beta. Instead, the relationship appears hump-shaped rather than linear, suggesting a possible quadratic structure, which is tested formally in Section @ref(extended-capm-specification).

## Security Market Line


```{r sml-estimation}
mean_market_premium <- capm_data %>%
  summarise(mean_excess_market = mean(excess_market, na.rm = TRUE)) %>%
  pull(mean_excess_market)

sml_model <- lm(mean_excess_return ~ portfolio_beta, data = portfolio_data)
sml_summary <- tidy(sml_model)

gamma0 <- sml_summary$estimate[1]
gamma1 <- sml_summary$estimate[2]
gamma0_p <- sml_summary$p.value[1]
gamma1_p <- sml_summary$p.value[2]
```
```{r sml-plot, fig.cap="Security Market Line estimated from portfolio data"}
portfolio_data %>%
  ggplot(aes(x = portfolio_beta, y = mean_excess_return)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "#d62728", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = expression(hat(beta)[p]),
    y = "Mean excess return"
  ) +
  theme_minimal(base_size = 12)
```
```{r sml-table}
sml_summary %>%
  mutate(term = c("$\\gamma_0$", "$\\gamma_1$")) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Estimate", "Std. Error", "t-statistic", "p-value"),
    caption = "Security Market Line regression results",
    booktabs = TRUE,
    align = "c",
    escape = FALSE
  )
```

The Security Market Line (SML) is estimated by regressing mean portfolio excess returns on portfolio betas:

\begin{equation}
r_{p_i} - r_f = \gamma_0 + \gamma_1 \beta_{p_{i}} + \varepsilon_{p_{i}}
\end{equation}

The estimated $\hat{\gamma}_0 =$ `r round(gamma0, 4)` (`r ifelse(gamma0_p < 0.05, "statistically significant", "not significant")` at 5%) and $\hat{\gamma}_1 =$ `r round(gamma1, 4)` (`r ifelse(gamma1_p < 0.05, "statistically significant", "not significant")` at 5%).
The observed mean market risk premium over the sample period is `r round(mean_market_premium, 4)`.

CAPM imposes two testable restrictions on the SML. First, $\gamma_0$ should be zero, since if portfolio excess returns are fully explained by systematic risk, no intercept should remain. 
Second, $\gamma_1$ should equal the observed market risk premium $\bar{r}_m - r_f$.

The intercept $\hat{\gamma}_0$ is `r ifelse(gamma0_p < 0.05, "statistically significant 
at the 5% level, which rejects the first CAPM restriction — some component of portfolio 
returns is not explained by market risk alone", "not statistically significant at the 5% 
level, consistent with the first CAPM restriction")`.

To test the second restriction, we assess whether $\gamma_1 = \bar{r}_m - r_f$ using a 
t-test:
\begin{equation}
t = \frac{\hat{\gamma}_1 - (\bar{r}_m - r_f)}{se(\hat{\gamma}_1)}
\end{equation}

```{r sml-gamma1-test}
gamma1_est <- sml_summary$estimate[2]
gamma1_se  <- sml_summary$std.error[2]
t_stat     <- (gamma1_est - mean_market_premium) / gamma1_se
p_val      <- 2 * pt(-abs(t_stat), df = nrow(portfolio_data) - 2)
```

The test yields $t =$ `r round(t_stat, 3)` (p-value = `r round(p_val, 3)`), so the 
hypothesis $\gamma_1 = \bar{r}_m - r_f$ is rejected at the 5% 
level. The estimated security market line exhibits a negative slope, indicating that higher-beta portfolios do not earn higher average returns in this sample. This contradicts the standard CAPM prediction and suggests that beta alone does not adequately explain the cross-sectional variation in returns

## Extended CAPM Specification {#extended-capm-specification}

Following Fama and MacBeth (1973) and Gibbons (1982), we test whether the linear CAPM specification is adequate by adding a squared beta term and a measure of idiosyncratic risk:

\begin{equation}
\bar{r}_{p} - r_f = \delta_1 + \delta_2 \hat{\beta}_p + \delta_3 \hat{\beta}_p^2 + \delta_4 \hat{\sigma}^2_{\varepsilon_p} + \varepsilon_p
\end{equation}

If CAPM holds, $\delta_3$ and $\delta_4$ should both be zero: $\delta_3 = 0$ tests for linearity, $\delta_4 = 0$ tests whether idiosyncratic risk is priced.
```{r extended-capm}
# Residual variance per stock from characteristic curves
resid_var <- capm_data %>%
  group_by(ticker) %>%
  do({
    m <- lm(excess_return ~ excess_market, data = .)
    tibble(sigma2_eps = var(resid(m)))
  }) %>%
  ungroup()

# Aggregate to portfolios
portfolio_sigma2 <- portfolio_assignments %>%
  left_join(resid_var, by = "ticker") %>%
  group_by(portfolio) %>%
  summarise(sigma2_eps = mean(sigma2_eps, na.rm = TRUE), .groups = "drop")

portfolio_data_ext <- portfolio_data %>%
  left_join(portfolio_sigma2, by = "portfolio") %>%
  mutate(beta2 = portfolio_beta^2)

extended_model <- lm(
  mean_excess_return ~ portfolio_beta + beta2 + sigma2_eps,
  data = portfolio_data_ext
)

tidy(extended_model) %>%
  mutate(term = c("$\\delta_1$", "$\\delta_2$", "$\\delta_3$", "$\\delta_4$")) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Estimate", "Std. Error", "t-statistic", "p-value"),
    caption = "Extended CAPM specification (Fama-MacBeth/Gibbons)",
    booktabs = TRUE,
    align = "c",
    escape = FALSE
  )
```

Table @ref(tab:extended-capm) reports the estimates of the extended CAPM specification. The coefficient $\delta_2$ is positive and statistically significant, indicating a positive relationship between beta and returns.

The coefficient $\delta_3$ is statistically significant, which suggests that the relationship between returns and beta is not linear. This contradicts the standard CAPM assumption.

The coefficient $\delta_4$ is not statistically significant, indicating that idiosyncratic risk is not priced. This result is consistent with CAPM.

The intercept $\delta_1$ is statistically significant, which suggests that average returns are not fully explained by beta.

Overall, the results provide only partial support for CAPM. The pricing of idiosyncratic risk is not supported, but the linearity assumption is violated and the intercept is significantly different from zero

# Structural Breaks and Asymmetry in CAPM

```{r break-setup}
# Determine split point: last 24 months
max_date <- max(capm_data$year_month)
break_date <- max_date %m-% months(48)

capm_data <- capm_data %>%
  mutate(
    period = if_else(year_month > break_date, "recent", "early"),
    post   = as.integer(year_month > break_date),
    excess_market_post = excess_market * post
  )
```

## Chow Tests on Individual Characteristic Curves

The CAPM characteristic curve is reformulated to allow for a change in $\beta$ over the last 48 months of the sample:

\begin{equation}
r_{j} - r_{f} = \alpha_{j} + \beta_{j}(r_{m} - r_{f}) + \delta_{j} \cdot D_t \cdot (r_{m} - r_{f}) + \varepsilon_{j}
\end{equation}

where $D_t = 1$ if $t$ falls in the last 48 months, and zero otherwise. Two variants of the Chow test are applied to each stock:

- **Chow structural break test**: tests whether all coefficients are identical across the two subperiods, using the full-sample residual sum of squares and the sum from two separate regressions.
- **Chow forecast (predictive) test**: tests whether the model estimated on the early period predicts returns in the recent period.

```{r chow-tests}
library(strucchange)

n_kept <- capm_data %>%
  group_by(ticker) %>%
  filter(sum(post == 0) >= 10, sum(post == 1) >= 5) %>%
  summarise() %>%
  nrow()

chow_results <- capm_data %>%
  group_by(ticker) %>%
  filter(sum(post == 0) >= 10, sum(post == 1) >= 5) %>%
  group_modify(~ {
    df <- .x %>% arrange(year_month)
    n_early  <- sum(df$post == 0)
    n_total  <- nrow(df)
    
    m_full  <- lm(excess_return ~ excess_market, data = df)
    m_early <- lm(excess_return ~ excess_market, data = filter(df, post == 0))
    m_late  <- lm(excess_return ~ excess_market, data = filter(df, post == 1))
    
    rss_full  <- sum(resid(m_full)^2)
    rss_early <- sum(resid(m_early)^2)
    rss_late  <- sum(resid(m_late)^2)
    
    k  <- 2
    df_full <- n_total - 2 * k
    
    # Chow structural break test
    F_chow <- ((rss_full - rss_early - rss_late) / k) /
              ((rss_early + rss_late) / df_full)
    p_chow <- pf(F_chow, k, df_full, lower.tail = FALSE)
    
    # Chow forecast test
    n_pred  <- sum(df$post == 1)
    df_fore <- n_early - k
    F_fore  <- ((rss_full - rss_early) / n_pred) /
               (rss_early / df_fore)
    p_fore  <- pf(F_fore, n_pred, df_fore, lower.tail = FALSE)
    
    tibble(F_chow = F_chow, p_chow = p_chow,
           F_fore = F_fore, p_fore = p_fore)
  }) %>%
  ungroup()
```
```{r chow-summary}
chow_summary <- chow_results %>%
  summarise(
    `Chow break — sig. at 5% [%]`    = mean(p_chow < 0.05) * 100,
    `Chow break — sig. at 1% [%]`    = mean(p_chow < 0.01) * 100,
    `Chow forecast — sig. at 5% [%]` = mean(p_fore < 0.05) * 100,
    `Chow forecast — sig. at 1% [%]` = mean(p_fore < 0.01) * 100
  ) 

chow_summary %>%
  kable(digits = 2, booktabs = TRUE, align = "c",
        caption = "Proportion of stocks with a significant structural break (\\%)") %>%
  kable_styling(latex_options = "scale_down")
```

At the 5% significance level, the Chow structural break test detects instability in about 13.9% of stocks, while at the 1% level this share falls to 4.6%. The Chow forecast test yields rejection rates of 14.1% at the 5% level and 9.1% at the 1% level.

Overall, most stocks do not show evidence of instability, but a non-negligible share does. Rejection rates are slightly higher in the forecast test, suggesting that out-of-sample fit is weaker than what parameter stability alone would indicate.

## Security Market Line by Subperiod
```{r sml-subperiod}
coef_characteristic_curve_struct <- capm_data %>%
  group_by(ticker, period) %>%
  do(tidy(lm(excess_return ~ excess_market, data = .))) %>%
  ungroup() %>%
  filter(term %in% c("(Intercept)", "excess_market")) %>%
  select(ticker, period, term, estimate, std.error, statistic, p.value) %>%
  pivot_wider(names_from = term, 
              values_from = c(estimate, std.error, statistic, p.value)) %>%
  rename(
    alpha     = `estimate_(Intercept)`,
    beta      = estimate_excess_market,
    alpha_se  = `std.error_(Intercept)`,
    beta_se   = std.error_excess_market,
    alpha_t   = `statistic_(Intercept)`,
    beta_t    = statistic_excess_market,
    alpha_p   = `p.value_(Intercept)`,
    beta_p    = p.value_excess_market
  )

portfolio_assignments_struct <- coef_characteristic_curve_struct %>%
  group_by(period) %>%
  arrange(beta, .by_group = TRUE) %>%
  mutate(portfolio = ntile(beta, n_portfolios)) %>%
  ungroup()
  
  
  portfolio_returns_struct <- capm_data %>%
  left_join(portfolio_assignments_struct %>% select(ticker, portfolio, period), by = c("ticker", "period")) %>%
  filter(!is.na(portfolio)) %>%
  group_by(portfolio, year_month, period) %>%
  summarise(
    portfolio_excess_return = mean(excess_return, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(portfolio, period) %>%
  summarise(
    mean_excess_return = mean(portfolio_excess_return, na.rm = TRUE),
    .groups = "drop"
  )

portfolio_betas_struct <- portfolio_assignments_struct %>%
  group_by(portfolio, period) %>%
  summarise(
    portfolio_beta = mean(beta, na.rm = TRUE),
    n_stocks = n(),
    .groups = "drop"
  )

portfolio_data_struct <- portfolio_betas_struct %>%
  left_join(portfolio_returns_struct, by = c("portfolio", "period"))


sml_split <- portfolio_data_struct %>%
  group_by(period) %>%
  group_modify(~ tidy(lm(mean_excess_return ~ portfolio_beta, data = .x))) %>%
  ungroup()
```
```{r sml-subperiod-plot, fig.cap="Security Market Line: early vs. recent subperiod"}
portfolio_data_struct %>%
  ggplot(aes(x = portfolio_beta, y = mean_excess_return, color = period)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_color_manual(values = c("early" = "steelblue", "recent" = "tomato"),
                     labels = c("early" = paste0("Early (up to ", format(break_date, "%b %Y"), ")"),
                                "recent" = paste0("Recent (last 48 months)"))) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = expression(hat(beta)[p]), y = "Mean excess return", color = "") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
```{r sml-subperiod-table}
sml_split %>%
  mutate(term = rep(c("$\\gamma_0$", "$\\gamma_1$"), 2)) %>%
  select(period, term, estimate, std.error, statistic, p.value) %>%
  kable(digits = 4, booktabs = TRUE, align = "c",
        escape = FALSE,
        col.names = c("Period", "Term", "Estimate", "Std. Error", "t-stat", "p-value"),
        caption = "SML estimates by subperiod")
```


Table \@ref(tab:sml-subperiod-table) and Figure \@ref(fig:sml-subperiod-plot) show the SML estimated separately for the early and recent subperiods. 

In the early period, the intercept ($\gamma_0$) is positive and significant, while the slope ($\gamma_1$) is negative and significant, implying a downward-sloping SML and a negative relation between beta and returns. This contradicts the CAPM prediction of a positive risk–return trade-off.

In the recent period, the intercept is small and only weakly significant, and the slope is close to zero and not significant. This indicates no meaningful relationship between beta and returns.

Overall, the estimated SMLs do not support the CAPM in either subperiod.

## Asymmetric Beta

The characteristic curve is reformulated to allow $\beta$ to differ depending on whether the market return is positive or negative:

\begin{equation}
r_{j} - r_{f} = \alpha_{j} + \beta_{j}^{+}(r_{m} - r_{f})^{+} + \beta_{j}^{-}(r_{m} - r_{f})^{-} + \varepsilon_{j}
\end{equation}

where $(r_{m} - r_{f})^{+} = \max(r_{m} - r_{f},\, 0)$ and $(r_{m} - r_{f})^{-} = \min(r_{m} - r_{f},\, 0)$.
```{r asymmetric-beta}
capm_data <- capm_data %>%
  mutate(
    market_up   = if_else(excess_market > 0, excess_market, 0),
    market_down = if_else(excess_market < 0, excess_market, 0)
  )

asym_coefs <- capm_data %>%
  group_by(ticker) %>%
  do(tidy(lm(excess_return ~ market_up + market_down, data = .))) %>%
  ungroup() %>%
  select(ticker, term, estimate, std.error, statistic, p.value) %>%
  pivot_wider(names_from = term,
              values_from = c(estimate, std.error, statistic, p.value)) %>%
  rename(
    beta_up     = estimate_market_up,
    beta_down   = estimate_market_down,
    beta_up_se  = std.error_market_up,
    beta_down_se = std.error_market_down,
    beta_up_t   = statistic_market_up,
    beta_down_t = statistic_market_down,
    beta_up_p   = p.value_market_up,
    beta_down_p = p.value_market_down
  ) %>%
  mutate(
    diff      = beta_up - beta_down,
    diff_se   = sqrt(beta_up_se^2 + beta_down_se^2),
    t_diff    = diff / diff_se,
    p_diff    = 2 * pt(-abs(t_diff), df = 58)
  )
```

For each stock, we test $H_0: \beta^{+} = \beta^{-}$ against $H_1: \beta^{+} \neq \beta^{-}$ at the 5% significance level using a t-test based on the asymptotic standard error of the difference.
```{r asymmetric-summary}
asym_coefs %>%
  summarise(
    `Mean beta+ ` = mean(beta_up,   na.rm = TRUE),
    `Mean beta-`  = mean(beta_down, na.rm = TRUE),
    `Mean diff`   = mean(diff,      na.rm = TRUE),
    `Sig. diff at 5% [%]` = mean(p_diff < 0.05, na.rm = TRUE) * 100
  ) %>%
  kable(digits = 3, booktabs = TRUE, align = "c",
        caption = "Asymmetric beta estimates: summary")
```

Table \@ref(tab:sml-updown-table) reports the results of the asymmetric beta estimates. The mean estimates of $\beta^{+}$ and $\beta^{-}$ are nearly identical (1.02 and 1.019), resulting in an average difference close to zero (0.001). This indicates no meaningful asymmetry in beta at the aggregate level.

At the individual stock level, 8.46% of stocks show a statistically significant difference between $\beta^{+}$ and $\beta^{-}$ at the 5% level. This suggests that asymmetry is present only for a limited subset of stocks rather than being a general pattern in the sample.

## Security Market Line in Up and Down Markets
```{r sml-updown}
capm_data <- capm_data %>%
  mutate(market_state = if_else(excess_market > 0, "up", "down"))

portfolio_returns_updown <- capm_data %>%
  left_join(portfolio_assignments %>% select(ticker, portfolio), by = "ticker") %>%
  filter(!is.na(portfolio)) %>%
  group_by(portfolio, market_state, year_month) %>%
  summarise(portfolio_excess_return = mean(excess_return, na.rm = TRUE),
            .groups = "drop") %>%
  group_by(portfolio, market_state) %>%
  summarise(mean_excess_return = mean(portfolio_excess_return, na.rm = TRUE),
            .groups = "drop")


portfolio_assignments_asym_up <- asym_coefs %>%
  arrange(beta_up, .by_group = TRUE) %>%
  mutate(portfolio = ntile(beta_up, n_portfolios))

portfolio_assignments_asym_down <- asym_coefs %>%
  arrange(beta_down, .by_group = TRUE) %>%
  mutate(portfolio = ntile(beta_down, n_portfolios)) 
    

portfolio_betas_asym_up <- portfolio_assignments_asym_up %>%
  group_by(portfolio) %>%
  summarise(
    portfolio_beta = mean(beta_up, na.rm = TRUE),
    n_stocks = n(),
    market_state = "up",
    .groups = "drop"
  )

portfolio_betas_asym_down <- portfolio_assignments_asym_down %>%
  group_by(portfolio) %>%
  summarise(
    portfolio_beta = mean(beta_down, na.rm = TRUE),
    n_stocks = n(),
    market_state = "down",
    .groups = "drop"
  )

portfolio_betas_asym <- bind_rows(
  portfolio_betas_asym_up,
  portfolio_betas_asym_down
)



portfolio_data_updown <- portfolio_betas_asym %>%
  left_join(portfolio_returns_updown, by = c("portfolio", "market_state"))

sml_updown <- portfolio_data_updown %>%
  group_by(market_state) %>%
  group_modify(~ tidy(lm(mean_excess_return ~ portfolio_beta, data = .x))) %>%
  ungroup()
```


```{r sml-updown-premium-test}
mean_market_premium_updown <- capm_data %>%
  group_by(market_state) %>%
  summarise(mean_excess_market = mean(excess_market, na.rm = TRUE), .groups = "drop")

mean_premium_down <- mean_market_premium_updown %>% 
  filter(market_state == "down") %>% pull(mean_excess_market)
mean_premium_up <- mean_market_premium_updown %>% 
  filter(market_state == "up") %>% pull(mean_excess_market)

# Extract gamma1 estimates and SEs for each state
gamma1_down    <- sml_updown %>% filter(market_state == "down", term == "portfolio_beta")
gamma1_up      <- sml_updown %>% filter(market_state == "up",   term == "portfolio_beta")

t_down <- (gamma1_down$estimate - mean_premium_down) / gamma1_down$std.error
t_up   <- (gamma1_up$estimate   - mean_premium_up)   / gamma1_up$std.error

p_down <- 2 * pt(-abs(t_down), df = nrow(portfolio_data_updown)/2 - 2)
p_up   <- 2 * pt(-abs(t_up),   df = nrow(portfolio_data_updown)/2 - 2)

tibble(
  `Market state` = c("Down", "Up"),
  `$\\hat{\\gamma}_1$` = c(gamma1_down$estimate, gamma1_up$estimate),
  `$\\bar{r}_m - r_f$` = c(mean_premium_down, mean_premium_up),
  `$t$-stat` = c(t_down, t_up),
  `$p$-value` = c(p_down, p_up)
) %>%
  kable(digits = 4, booktabs = TRUE, align = "c", escape = FALSE,
        caption = "Test of $\\gamma_1 = \\bar{r}_m - r_f$ by market state")
```



```{r sml-updown-plot, fig.cap="Security Market Line: up vs. down market months"}
portfolio_data_updown %>%
  ggplot(aes(x = portfolio_beta, y = mean_excess_return, color = market_state)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_color_manual(values = c("up" = "steelblue", "down" = "tomato")) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(x = expression(hat(beta)[p]), y = "Mean excess return", color = "Market state") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
```{r sml-updown-table}
sml_updown %>%
  mutate(term = rep(c("$\\gamma_0$", "$\\gamma_1$"), 2)) %>%
  select(market_state, term, estimate, std.error, statistic, p.value) %>%
  kable(digits = 4, booktabs = TRUE, align = "c",
        escape = FALSE,
        col.names = c("Market state", "Term", "Estimate", "Std. Error", "t-stat", "p-value"),
        caption = "SML estimates: up vs. down market months")
```

Figure @ref(fig:sml-updown-plot) and Table @ref(tab:sml-updown-table) show the SML estimates across market regimes, while Table @ref(tab:sml-updown-premium-test) reports the test of $\gamma_1 = \bar{r}_m - r_f$.

In down markets, the slope ($\gamma_1$) is negative and highly significant, indicating that higher-beta portfolios experience larger losses. The intercept ($\gamma_0$) is not statistically different from zero. The equality $\gamma_1 = \bar{r}_m - r_f$ is not rejected (p = 0.743), so the slope is consistent with the observed negative market premium.

In up markets, the slope is positive and significant but smaller than the market premium, and the equality $\gamma_1 = \bar{r}_m - r_f$ is strongly rejected (p < 0.001). The intercept is positive and significant, indicating that returns are not fully explained by beta.

In down markets, the CAPM restrictions are not rejected, as the slope is consistent with the market risk premium and the intercept is not statistically different from zero. In contrast, in up markets, the intercept is significantly different from zero and the slope is statistically not equal to the market risk premium, indicating a violation of the CAPM.
