---
title: "Analysis of Capital Asset Pricing Model (CAPM)"
author: 
  - "Filip Farkašovský"
  - "(Skupina 02, Tím 08)"
date: "`r format(Sys.Date(), '%d. %m. %Y')`"
lang: eng
output:
  bookdown::pdf_document2:
    highlight: monochrome
    fig_caption: true
---

# Abstract {-}

This study examines the Capital Asset Pricing Model (CAPM) using historical stock data from S&P 500 constituents. The analysis focuses on estimating beta coefficients for individual assets and evaluating whether systematic risk adequately explains differences in expected returns.

# Data Collection and Estimation of Beta Coefficients

## Data Loading and Preparation

```{r setup, include=FALSE}
rm(list = ls())
set.seed(1789)


knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE,
  fig.width = 5,
  fig.height = 3,
    out.width = "60%",
  fig.align = "center")

library(knitr)
library(tidyverse)
library(lubridate)
library(quantmod)
library(PerformanceAnalytics)
library(broom)
library(dplyr)
library(ggplot2)
library(car)
```

```{r data-loading, include=FALSE}
rm(list = ls())

# Load and preprocess stock dataset
unzip("SP500_Historical_Data.zip")
stocks <- read_csv("SP500_Historical_Data.csv") %>% 
    mutate(ticker = factor(Ticker),
           date = as.Date(Date),
           close = as.numeric(Close)) %>%
    arrange(ticker, date)

# Monthly prices
stocks <- stocks %>%
  mutate(year_month = floor_date(date, "month")) %>%
  arrange(ticker, year_month) %>%
  group_by(ticker, year_month) %>%
  summarise(price = last(close[!is.na(close)]), .groups = "drop") 

# Load and preprocess DTB3 dataset
risk_free_rate <- read_csv("DTB3.csv") %>%
  mutate(
    date = as.Date(observation_date),
    rf_annual = as.numeric(DTB3)
  ) %>%
  mutate(year_month = floor_date(date, "month")) %>%
  group_by(year_month) %>%
  summarise(rf_annual = last(rf_annual[!is.na(rf_annual)]), .groups = "drop") %>%
  mutate(rf_monthly = (rf_annual / 100) / 12)

# Join datasets and filter stock with few samples
capm_data <- stocks %>%
  left_join(risk_free_rate, by = "year_month")

n_before <- capm_data %>%
  summarise(n_tickers = n_distinct(ticker)) %>%
  pull(n_tickers)

# Filter stocks with insufficient number of observations
capm_data <- capm_data %>%
  group_by(ticker) %>%
  filter(!is.na(price)) %>%
  filter(n() >= 60) %>%
  ungroup()

n_after <- capm_data %>%
  summarise(n_tickers = n_distinct(ticker)) %>%
  pull(n_tickers)

# Logaritmical monthly revenue
capm_data <- capm_data %>%
  arrange(ticker, year_month) %>%
  group_by(ticker) %>%
  mutate(stock_return = log(price / lag(price))) %>%
  filter(!is.na(stock_return)) %>%
  ungroup()

# Obtain market returns for equal weighted stock portfolio
market_returns <- capm_data %>%
  group_by(year_month) %>%
  summarise(market_return = mean(stock_return), .groups = "drop")

capm_data <- capm_data %>%
  left_join(market_returns, by = "year_month")

# Add excess returns 
capm_data <- capm_data %>%
  mutate(
    excess_return = stock_return - rf_monthly,
    excess_market = market_return - rf_monthly
  )
```

We use a dataset containing historical daily stock price data for 472 current S&P 500 companies, covering the period from January 2000 to February 2026. This dataset is one of the most comprehensive freely available S&P 500 datasets on Kaggle, consisting of approximately 2.7 million observations over 26 years of U.S. equity market history. The dataset is available at: \url{https://www.kaggle.com/datasets/jacksaleeby/s-and-p500-historical-data}. To ensure sufficient time series length for reliable beta estimation, we retain only stocks with at least 60 monthly observations. After applying this filter, the dataset is reduced from `r n_before` stocks to `r n_after` stocks.

Monthly returns are computed from end-of-month prices as logarithic returns:

\begin{equation}
r_{j,t} = \ln\left(\frac{P_{j,t}}{P_{j,t-1}}\right)
\end{equation}

As the risk-free rate, we use the 3-Month Treasury Bill Secondary Market Rate (Discount Basis), which is a commonly used proxy for the risk-free rate in econometric studies. The data are obtained from the Federal Reserve Economic Data (FRED) database: \url{https://fred.stlouisfed.org/series/DTB3}.

The series is available at a daily frequency and in annualized percentage terms. It is converted to monthly frequency by taking the last observation of each month and transforming it into a monthly rate:

\begin{equation}
r_{f,t} = \frac{r_{f,t}^{(annual)}}{100 \cdot 12}.
\end{equation}

This transformation provides an approximation of the effective monthly risk-free rate.

## Visualization of Time Series
The time series in Figure @ref(fig:visualization) shows the evolution of the market return and the risk-free rate over time. The risk-free rate remains close to zero for most of the sample period and exhibits only limited variation, while the market return shows substantially higher volatility.


```{r visualization, fig.width=5, fig.height=3, out.width="70%", fig.cap="Time Series of Market Return and Risk-Free Rate"}
capm_data %>%
    ggplot(aes(x = year_month)) +
  geom_line(aes(y = market_return, color = "Market return"), linewidth = 0.8) +
  geom_line(aes(y = rf_monthly, color = "Risk-free rate"), linewidth = 0.8) +
  scale_color_manual(values = c("Market return" = "lightskyblue",
                                "Risk-free rate" = "tomato")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::percent_format()) +

  labs(
    x = "Year",
    y = "Monthly return",
    color = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
Table @ref(tab:summary-stat) reports summary statistics of monthly stock returns. The mean return is close to zero, while the dispersion is substantial, as indicated by the standard deviation.

```{r summary-stat}
capm_data %>%
  summarise(
    Mean = mean(stock_return, na.rm = TRUE),
    SD = sd(stock_return, na.rm = TRUE),
    Min = min(stock_return, na.rm = TRUE),
    Q1 = quantile(stock_return, 0.25, na.rm = TRUE),
    Median = median(stock_return, na.rm = TRUE),
    Q3 = quantile(stock_return, 0.75, na.rm = TRUE),
    Max = max(stock_return, na.rm = TRUE)
  ) %>%
  kable(digits = 4,booktabs = TRUE,
 caption = "Summary statistics of monthly stock returns")
```

Figure @ref(fig:stocks-dist) presents the distribution of monthly returns for a sample of stocks.

```{r stocks-dist, fig.cap="Distribution of month returns - 10 sample stocks"}
set.seed(1789)
sample_stocks <- sample(unique(capm_data$ticker), 10)


capm_data %>%
  filter(ticker %in% sample_stocks) %>%
  mutate(ticker = fct_reorder(ticker, stock_return, median)) %>%
  ggplot(aes(x = ticker, y = stock_return)) +
  geom_boxplot(fill = "lightskyblue", alpha = 0.6, outlier.size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "",
    x = NULL, y = "Month return"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7)
  )
```

## Estimate of betta coefficients of the characteristic curves

For each stock, the coefficients of the characteristic curve are estimated using the following model: 
\begin{equation}
r_{j} - r_{f} = \alpha_{j} + \beta_{j} (r_{m} - r_{f}) + \varepsilon_{j}
\end{equation}

```{r coeficient-estimation}
coef_characteristic_curve <- capm_data %>%
  group_by(ticker) %>%
  do(tidy(lm(excess_return ~ excess_market, data = .))) %>%
  ungroup() %>%
  filter(term %in% c("(Intercept)", "excess_market")) %>%
  select(ticker, term, estimate, std.error, statistic, p.value) %>%
  pivot_wider(names_from = term, 
              values_from = c(estimate, std.error, statistic, p.value)) %>%
  rename(
    alpha     = `estimate_(Intercept)`,
    beta      = estimate_excess_market,
    alpha_se  = `std.error_(Intercept)`,
    beta_se   = std.error_excess_market,
    alpha_t   = `statistic_(Intercept)`,
    beta_t    = statistic_excess_market,
    alpha_p   = `p.value_(Intercept)`,
    beta_p    = p.value_excess_market
  )

n_defensive_betas <- sum(coef_characteristic_curve$beta < 1)
n_agressive_betas <- sum(coef_characteristic_curve$beta > 1)
```

The CAPM beta coefficients are obtained from time-series regressions of excess stock returns on excess market returns. The results show substantial heterogeneity in systematic risk across assets.

As reported in Table \@ref(tab:tab-beta-summary), the average $\beta$ is close to one, which suggests that, on average, stocks have similar sensitivity to market movements as the market portfolio. However, individual estimates vary considerably. Some stocks have $\beta > 1$, indicating higher systematic risk, while others have $\beta < 1$ or values close to zero.

Table \@ref(tab:tab-beta-summary) also reports the proportion of statistically significant coefficients at different significance levels (5%, respectively 1%). These values represent the percentage of stocks for which the null hypothesis $\beta = 0$ is rejected at the corresponding level. A higher percentage implies stronger overall evidence that market returns explain individual stock returns.

The Security Characteristic Lines of selected stocks are shown in Figure \@ref(fig:scl-plot). 

```{r tab-beta-summary}
coef_characteristic_curve %>%
  summarise(
    Mean = mean(beta, na.rm = TRUE),
    Median = median(beta, na.rm = TRUE),
    Minimum = min(beta, na.rm = TRUE),
    Maximum = max(beta, na.rm = TRUE),
    `Significant (5%) [%]`  = mean(beta_p < 0.05, na.rm = TRUE) * 100,
    `Significant (1%) [%]`  = mean(beta_p < 0.01, na.rm = TRUE) * 100
  ) %>%
  kable(
    digits = 3,
    caption = "Summary statistics of estimated CAPM beta coefficients",
    booktabs = TRUE,
    align = "c"
  )
```

```{r scl-plot, fig.cap="Security Characteristic Lines of selected stocks"}

set.seed(123)

sample_stocks <- coef_characteristic_curve %>%
  slice_sample(n = 6) %>%
  pull(ticker)

capm_data %>%
  filter(ticker %in% sample_stocks) %>%
  ggplot(aes(x = excess_market, y = excess_return)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "#d62728") +
  facet_wrap(~ticker, scales = "free") +
  labs(
    x = "Excess market return",
    y = "Excess stock return"
  ) +
  theme_minimal()
```


## Testing hypotheses on the intercept of the CAPM model

For each stock, we test the hypothesis that the intercept (alpha) in the CAPM model is equal to zero:

\begin{equation}
H_0: \alpha_j = 0 \quad \text{vs.} \quad H_1: \alpha_j \neq 0
\end{equation}

The tests are conducted at the 5% significance level using t-tests based on the estimated regression coefficients.

The results are summarized in Table \@ref(tab:alpha-test-summary). The null hypothesis is rejected for a relatively small subset of stocks, indicating statistically significant abnormal returns. For these assets, the intercept differs significantly from zero, suggesting that the CAPM does not fully explain their returns.

Most stocks do not have an intercept that is significantly different from zero. This means that, for most assets, there is no evidence that they generate returns above what would be expected after accounting for market risk.


```{r alpha-test-summary, out.width="70%"}
library(gt)

coef_characteristic_curve %>%
  mutate(`Significant on 5%` = alpha_p < 0.05) %>%
  group_by(`Significant on 5%`) %>%
  summarise(count = n()) %>%
  mutate(percentage = count / sum(count) * 100) %>%
  kable(
    digits = 2,
    booktabs = TRUE,
    align = "c",
    caption = "CAPM intercept significance (5\\% level)"
  )
```


# Testing CAPM Validity

## Portfolio Formation and Beta Aggregation

Stocks are sorted by their estimated $\beta$ coefficients and assigned to 20 portfolios of roughly equal size. Each portfolio is equally weighted, so the portfolio beta is the arithmetic mean of individual betas. Portfolio excess returns are computed as the time-series average of equally weighted stock excess returns over the full sample period.
```{r portfolio-formation}
n_portfolios <- 20

portfolio_assignments <- coef_characteristic_curve %>%
  arrange(beta) %>%
  mutate(portfolio = ntile(beta, n_portfolios))

portfolio_returns <- capm_data %>%
  left_join(portfolio_assignments %>% select(ticker, portfolio), by = "ticker") %>%
  filter(!is.na(portfolio)) %>%
  group_by(portfolio, year_month) %>%
  summarise(
    portfolio_excess_return = mean(excess_return, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(portfolio) %>%
  summarise(
    mean_excess_return = mean(portfolio_excess_return, na.rm = TRUE),
    .groups = "drop"
  )

portfolio_betas <- portfolio_assignments %>%
  group_by(portfolio) %>%
  summarise(
    portfolio_beta = mean(beta, na.rm = TRUE),
    n_stocks = n(),
    .groups = "drop"
  )

portfolio_data <- portfolio_betas %>%
  left_join(portfolio_returns, by = "portfolio")
```

Table @ref(tab:portfolio-table) reveals a pattern inconsistent with CAPM. While portfolios are clearly ordered by beta, mean excess returns show no systematic increase with beta. As visible in Figure @ref(fig:sml-plot), the relationship between beta and mean excess return appears hump-shaped rather than linear, suggesting a possible quadratic structure — this is tested formally in Section @ref(extended-capm-specification).
```{r portfolio-table}
portfolio_data %>%
  kable(
    digits = 4,
    col.names = c("Portfolio", "Beta", "N Stocks", "Mean Excess Return"),
    caption = "Portfolio betas and mean excess returns",
    booktabs = TRUE,
    align = "c"
  )
```

## Security Market Line

The Security Market Line (SML) is estimated by regressing mean portfolio excess returns on portfolio betas:

\begin{equation}
r_{p_i} - r_f = \gamma_0 + \gamma_1 \beta_{p_{i}} + \varepsilon_{p_{i}}
\end{equation}

If CAPM holds, $\gamma_0$ should be statistically indistinguishable from zero and $\gamma_1$ should equal the observed mean market risk premium $\bar{r}_m - r_f$.
```{r sml-estimation}
mean_market_premium <- capm_data %>%
  summarise(mean_excess_market = mean(excess_market, na.rm = TRUE)) %>%
  pull(mean_excess_market)

sml_model <- lm(mean_excess_return ~ portfolio_beta, data = portfolio_data)
sml_summary <- tidy(sml_model)

gamma0 <- sml_summary$estimate[1]
gamma1 <- sml_summary$estimate[2]
gamma0_p <- sml_summary$p.value[1]
gamma1_p <- sml_summary$p.value[2]
```
```{r sml-plot, fig.cap="Security Market Line estimated from portfolio data"}
portfolio_data %>%
  ggplot(aes(x = portfolio_beta, y = mean_excess_return)) +
  geom_point(size = 2.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "#d62728", linewidth = 0.8) +
  geom_abline(
    intercept = 0,
    slope = mean_market_premium,
    linetype = "dashed", color = "grey40"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = expression(hat(beta)[p]),
    y = "Mean excess return"
  ) +
  theme_minimal(base_size = 12)
```
```{r sml-table}
sml_summary %>%
  mutate(term = c("$\\gamma_0$", "$\\gamma_1$")) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Estimate", "Std. Error", "t-statistic", "p-value"),
    caption = "Security Market Line regression results",
    booktabs = TRUE,
    align = "c",
    escape = FALSE
  )
```

The estimated $\hat{\gamma}_0 =$ `r round(gamma0, 4)` (`r ifelse(gamma0_p < 0.05, "statistically significant", "not significant")` at 5%) and $\hat{\gamma}_1 =$ `r round(gamma1, 4)` (`r ifelse(gamma1_p < 0.05, "statistically significant", "not significant")` at 5%). The observed mean market risk premium is `r round(mean_market_premium, 4)`.

These results are in line with typical empirical findings: $\gamma_1$ tends to be below the observed market premium, meaning defensive stocks earn more than CAPM predicts and aggressive stocks earn less. A non-zero $\gamma_0$ would indicate the presence of returns not explained by market risk alone, which works against CAPM.

The estimated $\hat{\gamma}_0 =$ `r round(gamma0, 4)` and $\hat{\gamma}_1 =$ `r round(gamma1, 4)`. 
The observed mean market risk premium over the sample period is `r round(mean_market_premium, 4)`.

CAPM imposes two testable restrictions on the SML. First, $\gamma_0$ should be zero — if 
portfolio excess returns are fully explained by systematic risk, no intercept should remain. 
Second, $\gamma_1$ should equal the observed market risk premium $\bar{r}_m - r_f$.

The intercept $\hat{\gamma}_0$ is `r ifelse(gamma0_p < 0.05, "statistically significant 
at the 5% level, which rejects the first CAPM restriction — some component of portfolio 
returns is not explained by market risk alone", "not statistically significant at the 5% 
level, consistent with the first CAPM restriction")`.

To test the second restriction, we assess whether $\gamma_1 = \bar{r}_m - r_f$ using a 
t-test:

$$t = \frac{\hat{\gamma}_1 - (\bar{r}_m - r_f)}{se(\hat{\gamma}_1)}$$
```{r sml-gamma1-test}
gamma1_est <- sml_summary$estimate[2]
gamma1_se  <- sml_summary$std.error[2]
t_stat     <- (gamma1_est - mean_market_premium) / gamma1_se
p_val      <- 2 * pt(-abs(t_stat), df = nrow(portfolio_data) - 2)
```

The test yields $t =$ `r round(t_stat, 3)` (p-value = `r round(p_val, 3)`), so the 
hypothesis $\gamma_1 = \bar{r}_m - r_f$ is `r ifelse(p_val < 0.05, "rejected at the 5% 
level. The estimated slope is below the observed market premium, consistent with the 
well-documented finding that defensive stocks earn more than CAPM predicts and aggressive 
stocks earn less", "not rejected at the 5% level, meaning the slope is statistically 
consistent with the observed market risk premium")`.

Taken together, these results `r ifelse(gamma0_p < 0.05 | p_val < 0.05, "do not support", 
"are consistent with")` the CAPM restrictions on the SML.

## Extended CAPM Specification {#extended-capm-specification}

Following Fama and MacBeth (1973) and Gibbons (1982), we test whether the linear CAPM specification is adequate by adding a squared beta term and a measure of idiosyncratic risk:

\begin{equation}
\bar{r}_{p} - r_f = \delta_1 + \delta_2 \hat{\beta}_p + \delta_3 \hat{\beta}_p^2 + \delta_4 \hat{\sigma}^2_{\varepsilon_p} + \varepsilon_p
\end{equation}

If CAPM holds, $\delta_3$ and $\delta_4$ should both be zero: $\delta_3 = 0$ tests for linearity, $\delta_4 = 0$ tests whether idiosyncratic risk is priced.
```{r extended-capm}
# Residual variance per stock from characteristic curves
resid_var <- capm_data %>%
  group_by(ticker) %>%
  do({
    m <- lm(excess_return ~ excess_market, data = .)
    tibble(sigma2_eps = var(resid(m)))
  }) %>%
  ungroup()

# Aggregate to portfolios
portfolio_sigma2 <- portfolio_assignments %>%
  left_join(resid_var, by = "ticker") %>%
  group_by(portfolio) %>%
  summarise(sigma2_eps = mean(sigma2_eps, na.rm = TRUE), .groups = "drop")

portfolio_data_ext <- portfolio_data %>%
  left_join(portfolio_sigma2, by = "portfolio") %>%
  mutate(beta2 = portfolio_beta^2)

extended_model <- lm(
  mean_excess_return ~ portfolio_beta + beta2 + sigma2_eps,
  data = portfolio_data_ext
)

tidy(extended_model) %>%
  mutate(term = c("$\\delta_1$", "$\\delta_2$", "$\\delta_3$", "$\\delta_4$")) %>%
  kable(
    digits = 4,
    col.names = c("Term", "Estimate", "Std. Error", "t-statistic", "p-value"),
    caption = "Extended CAPM specification (Fama-MacBeth/Gibbons)",
    booktabs = TRUE,
    align = "c",
    escape = FALSE
  )
```

The significance of $\delta_3$ indicates whether the return-beta relationship deviates from linearity. A significant $\delta_4$ would suggest that idiosyncratic risk carries a risk premium, which CAPM does not allow for. Interpret the estimates in the context of the p-values reported in Table \@ref(tab:extended-capm).
